Highly Available Transactions: Virtues and Limitations

### 2013-03-29 writeup/conclusion.tex

Fundamentally, high availability represents a new constraint for
database designs. While the CAP Theorem has had a large impact on
distributed data storage design, our knowledge of database systems has
not kept pace. The key distributed systems challenge of designing
distributed algorithms and architectures for partial failure and
network asynchrony is largely unaddressed in the literature on
transaction processing systems. We believe that, in order to fully
realize the potential of HAT systems, further rearchitecture and study
of existing database designs as well as better understanding of
application semantics and their relationship to distributed
consistency will be necessary.

>>>>

Fundamentally, high availability represents a new constraint for
database designs. While the CAP Theorem has had a large impact on
highly available distributed data storage design, our knowledge of
database systems has not kept pace. The key distributed systems
challenge of designing distributed algorithms and architectures for
partial failure and network asynchrony is largely unaddressed in the
literature on transaction processing systems. To fully realizing the
potential of HAT systems will require further rearchitecture and study
of existing database designs as well as better understanding of
application semantics and their relationship to distributed
consistency.

### 2013-03-29 writeup/evaluation.tex

\section{Analysis}
\label{sec:evaluation}

@@

\begin{myenumerate}
\item We revisit traditional database concurrency control with a focus
  on coordination costs and on high availability.
\item We examine the properties required by a realistic OLTP
  application patterned on TPC-C.
\item We perform a brief experimental evaluation of HAT versus non-HAT
  properties on public cloud infrastructure.
\end{myenumerate}

@@

\vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
order on transactions, all existing algorithms for achieving
serializability of general-purpose read-write transactions in a
distributed setting~\cite{bernstein-book} may require at least one
round trip time (RTT) before committing. As an example, traditional
two-phase locking for a transaction of length $T$ requires $T$
\texttt{lock} operations and at least one unlock operation.  Each of
these operations requires coordination, either with other database
replicas or with a lock service. If the coordination mechanism for
locking is unavailable, transactions cannot make progress. Similarly,
optimistic concurrency control requires contacting a validation
service, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Multi-version concurrency control fares somewhat better but
must still be checked for write conflicts.

@@

\vspace{.5em}\noindent\textbf{Non-serializablity} Perhaps
interestingly, many existing distributed implementations of weak
isolation are not highly available. Lock-based proposals such as those
used to provide weak isolation in Gray's original
proposal~\cite{gray-isolation} do not degrade gracefully in the
presence of partial failure. However, lock-based protocols \textit{do}
provide recency guarantees. While multi-versioned storage systems
allow for a variety of transactional guarantees, we have not seen
proposals for truly weak isolation (e.g., non-``tentative update''
schemes) in this context. Chan and Gray's read-only transactions
provide read-only transactions with item-cut isolation, causal
consistency, and transactional atomicity but are unavailable in the
presence of coordinator failure~\cite{readonly}, similar to read-only
and write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
Swift~\cite{swift}, Brantner's S3 database~\cite{kraska-s3}, and
Bayou~\cite{sessionguarantees} can all provide variants of the above
model with high availability, but none provide this HAT functionality
without modification. Similarly, the MDCC~\cite{mdcc} protocol offers
Read Committed isolation with Lost Update avoidance but is similarly
unavailable due to its reliance on preventing write conflicts.

@@

Thus far, we have largely ignored the question of which applications
require semantics which are unavailable to HATs. As we showed in
Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
inability to prevent Lost Update, Write Skew, and provide recency
bounds. In this Section, we attempt to understand when these
guarantees matter both abstractly and by the use of a representative
customer order transaction application patterned off of TPC-C.

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent work on
the CALM Theorem~\cite{calm}, Lattice-based state
update~\cite{blooml}, and Commutative and Replicated Data
Types~\cite{crdt} demonstrates that, if updates logically commute,
then they can be performed in different orders at different locations
in a distributed system. Accordingly, as long as all writes are
delivered to all replicas, then a system executing monotonic logic
with commutative operators should not suffer from application-level
consistency anomalies as a result of Lost Update or Write Skew
anomalies. However, applications with non-monotonic state mutation
will not, in general, be able to maintain application-level
consistency constraints. Similarly, applications requiring write
visibility between clients should opt for unavailability instead of
possibly missing a write visibility deadline.

\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient
result describing exactly which application logics can be achieved
with high availability and to make these requirements more concrete,
we consider an example application: the TPC-C benchmark. Our goal is
\textit{not} to provide a performance analysis, as any HAT
configuration will be non-compliant. Rather, we examine the five
queries contained in the benchmarks---ostensibly representative of a
standard transaction processing application---to understand which are
achievable in a HAT context. Surprisingly, four of five can be
executed with HATs, while the fifth potentially requires unavailablity.

TPC-C consists of five transactions, capturing operation of a
real-world warehouse, including sales, payments, and deliveries. Two
transactions---Order-Status and Stock-Level---are read-only and can be
executed safely under HATs. Clients may read stale data, but this does
not violate TPC-C requirements and clients will read their writes if
they are sticky-available. Another transaction type, Payment, updates
running balances for warehouses, districts, and customer records as
well as providing an audit trail. The transaction is increment-only,
so all balance increase operations can commute, and TA allows the
maintainence of any foreign-key constraints (e.g.,
\texttt{UPDATE/DELETE CASCADE}).

While three out of five transactions have been easily achievable with
HATs, the remaining two transactions---New-Order and Delivery---are
not so simple. The New-Order transaction places an order for a
variable quantity of data items, updating warehouse stock as
needed. It selects a sales district, assigns the order an ID number,
adjusts the remaining warehouse stock, and writes a placeholder entry
for the pending order. The Delivery transaction represents the
fulfillment of a New-Order: it deletes the order from the pending
list, updates the customer's balance, updates the order's carrier ID
and delivery time, and updates the customer balance.

There are three challenges that only allow HATs to place non-compliant
New-Orders and to fail on Payments. First, we must assign each new
order an ID number. We can assign a unique number by, say,
concatenating the client ID and a timestamp. However, the TPC-C
specification requires order numbers to be assigned sequentially,
which requires Lost Update. Accordingly, HATs cannot provide compliant
TPC-C execution but can maintain uniqueness constraints. Second, the
New-Order transaction decrements inventory counts: what if the count
becomes negative? Fortunately, TPC-C New-Order restocks the inventory
count (increments by 91) if it would become negative as the result of
placing an order. This means that, even in the presence of concurrent
New-Orders, an item's stock will never fall below zero. This is TPC-C
compliant, but a HAT system might end up with substantially more stock
than in a single-site implementation. Third, delivery events are
non-monotonic---deleting a pending New-Order transaction---and should
be idempotent in order to avoid billing a customer twice; this implies
a need to prevent Lost Update. We can avoid this issue by moving the
non-monotonicity to the real world---the carrier that picks up the
package for an order can ensure that she is the only carrier who has
done so---but cannot provide a correct execution with HATs alone.

Throughout execution, TPC-C also requires the maintainence of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update both tables atomically via
transactional atomicity, can be easily maintained. Consistency
Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
applying updates atomically across tables. Consistency Conditions 2
and 3 (3.3.2.2-3) concern order ID assignment and are more
problematic, as we have discussed.

In summary, many---but not all---TPC-C transactions are well-served by
HATs. The two problematic transactions---New-Order and Payment---rely
on non-monotonic state update. The former transaction can be modified
to ensure ID uniqueness at the expense of sequential per-district ID
ordering, while the latter is an inherently non-monotonic action
requiring external compensation or stronger consistency protocols. We
expect that, especially for read-dominated workloads found in many
online services, HAT guarantees will suffice for many
applications. Finally, while TPC-C is not subject to write skew
anomalies, we note that many TPC-E isolation tests, such as one that
requires simultaneously modifying a product description and its image,
are satisfied by HATs.

>>>>

\section{Application and System Analysis}
\label{sec:evaluation}

@@

\begin{myenumerate}
\item We revisit traditional database concurrency control with a focus
  on coordination costs and on high availability.
\item We examine the properties required by a realistic OLTP
  application patterned on the TPC-C benchmark.
\item We perform a brief experimental evaluation of HAT versus non-HAT
  properties on public cloud infrastructure.
\end{myenumerate}

@@

\vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
order on transactions, all existing algorithms for achieving
serializability of general-purpose read-write transactions in a
distributed setting~\cite{bernstein-book} may require at least one
round trip time (RTT) before committing. As an example, traditional
two-phase locking for a transaction of length $T$ requires $T$
\texttt{lock} operations and at least one \texttt{unlock} operation.
Each of these operations requires coordination, either with other
database replicas or with a lock service. If the coordination
mechanism for locking is unavailable, transactions cannot
commit. Similarly, optimistic concurrency control requires contacting
a validation service, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
fares somewhat better but still requires checking for update
conflicts.

@@

\vspace{.5em}\noindent\textbf{Non-serializablity} Perhaps
interestingly, many existing distributed implementations of weak
isolation are not highly available. Lock-based proposals such as those
used to provide weak isolation in Gray's original
proposal~\cite{gray-isolation} do not degrade gracefully in the
presence of partial failure. (Note, however, that lock-based protocols
\textit{do} offer the benefit of recency guarantees.) While
multi-versioned storage systems allow for a variety of transactional
guarantees, we have not seen proposals for truly weak isolation (e.g.,
non-``tentative update'' schemes) in this context. Chan and Gray's
read-only transactions provide read-only transactions with item-cut
isolation, causal consistency, and transactional atomicity but are
unavailable in the presence of coordinator failure~\cite{readonly},
similar to read-only and write-only transactions more recently
proposed by Eiger~\cite{eiger}.  Bolt-on causal
consistency~\cite{bolton}, Brantner's S3 database~\cite{kraska-s3},
and Bayou~\cite{sessionguarantees} can all provide variants of the
above model with high availability, but none provide this HAT
functionality without substantial modification; Swift~\cite{swift} is
closest to providing full HAT operation in a sticky
context. Similarly, the MDCC~\cite{mdcc} protocol offers Read
Committed isolation with Lost Update avoidance but is similarly
unavailable due to its reliance on preventing write conflicts.

@@

Thus far, we have largely ignored the question of which applications
require semantics which are unavailable to HATs. As we showed in
Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
inability to prevent Lost Update, Write Skew, and provide recency
bounds. In this section, we attempt to understand when these
guarantees matter both abstractly and by the use of a representative
customer order transaction application patterned off of the TPC-C
benchmark~\cite{tpcc}.

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
work on the CALM Theorem~\cite{calm}, Lattice-based state
update~\cite{blooml}, and Commutative and Replicated Data
Types~\cite{crdt} demonstrates that, if updates logically commute,
then they can often be safely performed in different orders at
different locations in a distributed system. Accordingly, as long as
all writes are delivered to all replicas, then a system executing
monotonic logic with commutative operators may not suffer from
application-level consistency anomalies as a result of Lost Update or
Write Skew anomalies. However, applications with non-monotonic state
mutation will not, in general, be able to maintain application-level
consistency constraints. Similarly, applications requiring write
visibility between clients should opt for unavailability instead of
possibly missing a write visibility deadline.

\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient
result describing exactly which application logics can be achieved
with high availability and to make these requirements more concrete,
we consider an example application: the TPC-C benchmark. Our goal is
\textit{not} to provide a performance analysis, as a HAT
configuration will be non-compliant. Rather, we examine the five
queries contained in the benchmarks---ostensibly representative of a
standard transaction processing application---to understand which are
achievable in a HAT context. Surprisingly, four of five can be
executed with HATs, while the fifth potentially requires unavailablity.

TPC-C consists of five transactions, capturing operation of a
real-world warehouse, including sales, payments, and deliveries. Two
transactions---\textit{Order-Status} and \textit{Stock-Level}---are
read-only and can be executed safely under HATs. Clients may read
stale data, but this does not violate TPC-C requirements and clients
will read their writes if they are sticky-available. Another
transaction type, \textit{Payment}, updates running balances for
warehouses, districts, and customer records as well as providing an
audit trail. The transaction is increment-only, so all balance
increase operations can commute, and TA allows the maintainence of any
foreign-key constraints (e.g., \texttt{UPDATE/DELETE CASCADE}).

While three out of five transactions have been easily achievable with
HATs, the remaining two transactions---\textit{New-Order} and
\textit{Delivery}---are not so simple. The New-Order transaction
places an order for a variable quantity of data items, updating
warehouse stock as needed. It selects a sales district, assigns the
order an ID number, adjusts the remaining warehouse stock, and writes
a placeholder entry for the pending order. The Delivery transaction
represents the fulfillment of a New-Order: it deletes the order from
the pending list, updates the customer's balance, updates the order's
carrier ID and delivery time, and updates the customer balance.

There are three challenges that only allow HATs to place non-compliant
New-Orders and to fail on Payments. First, New-Order requires a unique
ID number for each order. We can create a unique number by, say,
concatenating the client ID and a timestamp. However, the TPC-C
specification requires order numbers to be \textit{sequentially}
assigned, which requires Lost Update. Accordingly, HATs cannot provide
compliant TPC-C execution but can maintain uniqueness
constraints. Second, the New-Order transaction decrements inventory
counts: what if the count becomes negative? Fortunately, TPC-C
New-Order restocks each item's inventory count (increments by 91) if
it would become negative as the result of placing an order. This means
that, even in the presence of concurrent New-Orders, an item's stock
will never fall below zero. This is TPC-C compliant, but a HAT system
might end up with substantially more stock than in a single-site
implementation. Third, delivery events are non-monotonic---deleting a
pending New-Order transaction---and should be idempotent in order to
avoid billing a customer twice; this implies a need to prevent Lost
Update. We can avoid this issue by moving the non-monotonicity to the
real world---the carrier that picks up the package for an order can
ensure that she is the only carrier who has done so---but cannot
provide a correct execution with HATs alone.

Throughout execution, TPC-C also requires the maintainence of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update rows in both tables atomically
via transactional atomicity, can be easily maintained. Consistency
Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
applying updates atomically across tables. Consistency Conditions 2
and 3 (3.3.2.2-3) concern order ID assignment and are more
problematic, as we have discussed.

In summary, many---but not all---TPC-C transactions are well-served by
HATs. The two problematic transactions---New-Order and Payment---rely
on non-monotonic state update. The former transaction can be modified
to ensure ID uniqueness at the expense of sequential per-district ID
ordering, while the latter is an inherently non-monotonic action
requiring external compensation or stronger consistency protocols. We
expect that, especially for read-dominated workloads found in many
online services, HAT guarantees will suffice for many application
transactions. Finally, while TPC-C is not subject to multi-key
anomalies, we note that many TPC-E isolation tests, such as one that
requires simultaneously modifying a product description and its image,
are also passable using HATs.

### 2013-03-29 writeup/futurework.tex

\textbf{What is the space of consistency and isolation properties?} In
this work, we have taxonomized several well-defined and
well-documented consistency properties. This is useful as it
contextualizes decades of existing database and distributed systems
design and algorithms but does not provide guidance as to what
consistency models are the ``strongest''---or admit the smallest set
of executions---within each availability class. A recent report from
UT Austin shows that no model stronger than causal consistency can be
achieved in a sticky highly available, one-way convergent
system~\cite{cac}, but there are infinite highly available and sticky
highly available models to consider. For example, always returning the
value $1$ in response to reads is incomparable with (i.e., neither
stronger nor weaker than) causal consistency: returning $1$ does not
respect causality, and causal consistency returns values other than
$1$. Always returning the value $2$ is similarly incomparable, and, as
an example, we can enumerate models that return all integers and real
numbers--an uncountably infinite set of models. Given these infinite
models, are there meaningful properties that delineate the boundaries
between classes of highly available systems? Is there a
``meta-property'' that exactly characterizes what properties are
achievable with high availability or sticky high availability? This is
an area of ongoing research within the Berkeley HAT project.

>>>>

\textbf{What are the limits of HAT systems, and how are they defined?}
In this work, we have taxonomized several well-defined and
well-documented consistency properties. This is useful as it
contextualizes decades of existing database and distributed systems
design and algorithms but does not provide guidance as to what
consistency models are the ``strongest''---or admit the smallest set
of executions---within each availability class. A recent report from
UT Austin shows that no model stronger than causal consistency can be
achieved in a sticky highly available, one-way convergent
system~\cite{cac}, but there are infinite highly available and sticky
highly available models to consider. For example, always returning the
value $1$ in response to reads is incomparable with (i.e., neither
stronger nor weaker than) causal consistency: returning $1$ does not
respect causality, and causal consistency returns values other than
$1$. Always returning the value $2$ is similarly incomparable, and, as
an example, we can enumerate models that return all integers and real
numbers--an uncountably infinite set of models. Given these infinite
models, are there meaningful properties that delineate the boundaries
between classes of highly available systems? Is there a
``meta-property'' that exactly characterizes what properties are
achievable with high availability or sticky high availability? This is
an area of ongoing research within the Berkeley HAT project.

### 2013-03-29 writeup/hat-vldb.tex

\begin{abstract}
Given the presence of network latency and the possibility of partial
failure and network partitions, many modern distributed systems eschew
transactional functionality, which provides strong semantic guarantees
on groups of multiple operations over multiple data items. In this
work, we consider the problem of providing transactional guarantees
that are highly available, or do not suffer from unavailability during
system partitions or typically incur high network latency.  We
introduce a new taxonomy of highly available algorithms and analyze a
combination of ACID and distributed consistency guarantees to identify
which can and cannot be achieved by Highly Available Transaction (HAT)
systems. In doing so, we analytically and experimentally quantify the
benefit of providing high availability as well as the associated
semantic costs.
\end{abstract}

>>>>

\begin{abstract}
Given the presence of network latency and the possibility of partial
failure and network partitions, many modern distributed systems eschew
transactional functionality, which provides strong semantic guarantees
for groups of multiple operations over multiple data items. In this
work, we consider the problem of providing transactional guarantees
that are highly available, or do not suffer from unavailability during
system partitions or typically incur high network latency.  We
introduce a new taxonomy of highly available algorithms and analyze a
combination of ACID and distributed consistency guarantees to identify
which can and cannot be achieved by Highly Available Transaction (HAT)
systems. In doing so, we analytically and experimentally quantify the
benefit of providing high availability as well as the associated
semantic costs.
\end{abstract}
